### 2015-06-20 chapters/introduction.tex
adding two new unicode pittfalls

The (default) \textsc{grapheme clusters} are locale-independent graphemes,
i.e.~they always apply when a particular combination of characters occurs
independent of the writing system in which they are used. These character
combinations are defined in the Unicode Standard as functioning as one
\textsc{text element}.\footnote{The Unicode Glossary defines text element as:
``A minimum unit of text in relation to a particular text process, in the
context of a given writing system. In general, the mapping between text elements
and code points is many-to-many.''} The simplest example of a grapheme cluster
is a base character followed by a letter modifier character. For example, the
sequence <n> + < ̃> (i.e.~\textsc{latin small letter n} at \uni{006E}, followed
by \textsc{combining tilde} at \uni{0303}) combines visually into <ñ>, a
user-perceived character in writing systems like that of Spanish. So, what the
user perceives as a single character actually involves a multi-code-point
sequence. Note that this specific sequence can also be represented with a single
so-called \textsc{precomposed} code point, \textsc{latin small letter n with
tilde} at \uni{00F1}, but this is not the case for all multi-code point
character sequences. The problem that there multiple encodings possible for the
same text element has been acknowledged early on in the Unicode Standard
(e.g.~for <ñ>, the sequence \uni{006E} \uni{0303} should in all situations be
treated identically to the precomposed \uni{00F1}), and a system of
\textsc{canonical equivalence} is available for such situations. Basically, the
Unicode Standard offers different kind of normalizations to either decompose all
precomposed characters (called \textsc{NFD}, \textsc{normalization form
canonical decomposition}), or compose as much as possible combinations (called
\textsc{NFC}, \textsc{normalization form canonical composition}). In current
practice of software development, NFC seems to be preferred in most situations
and is widely proposed as the preferred canonical form.

>>>>

The (default) \textsc{grapheme clusters} are locale-independent graphemes,
i.e.~they always apply when a particular combination of characters occurs
independent of the writing system in which they are used. These character
combinations are defined in the Unicode Standard as functioning as one
\textsc{text element}.\footnote{The Unicode Glossary defines text element as:
``A minimum unit of text in relation to a particular text process, in the
context of a given writing system. In general, the mapping between text elements
and code points is many-to-many.''} The simplest example of a grapheme cluster
is a base character followed by a letter modifier character. For example, the
sequence <n> + <\dia{0303}> (i.e.~\textsc{latin small letter n} at \uni{006E}, followed
by \textsc{combining tilde} at \uni{0303}) combines visually into <ñ>, a
user-perceived character in writing systems like that of Spanish. So, what the
user perceives as a single character actually involves a multi-code-point
sequence. Note that this specific sequence can also be represented with a single
so-called \textsc{precomposed} code point, \textsc{latin small letter n with
tilde} at \uni{00F1}, but this is not the case for all multi-code point
character sequences. The problem that there multiple encodings possible for the
same text element has been acknowledged early on in the Unicode Standard
(e.g.~for <ñ>, the sequence \uni{006E} \uni{0303} should in all situations be
treated identically to the precomposed \uni{00F1}), and a system of
\textsc{canonical equivalence} is available for such situations. Basically, the
Unicode Standard offers different kind of normalizations to either decompose all
precomposed characters (called \textsc{NFD}, \textsc{normalization form
canonical decomposition}), or compose as much as possible combinations (called
\textsc{NFC}, \textsc{normalization form canonical composition}). In current
practice of software development, NFC seems to be preferred in most situations
and is widely proposed as the preferred canonical form.

### 2015-06-20 chapters/ipa_meets_unicode.tex
adding two new unicode pittfalls

The International Phonetic Alphabet (IPA) is a common standard in linguistics to
transcribe sounds of spoken language into Latin-based characters. Although IPA
is reasonably easily adhered to with pen and paper, it is not trivial to encode
IPA characters electronically. Similar to the previous chapter, this chapter
will discuss various pitfalls with the Unicode encoding of the IPA.\@ We will
specifically refer to the 2005 version of the IPA \citep{IPA2005} and the 7.0 
version of Unicode \citet{Unicode2014}.

@@

Tie bars are a special type of character in the sense that they do not belong to
a segment, but bind two graphemes together. This actually turns out to be rather
different from the Unicode conceptions, see
Section~\ref{pitfall-different-notions-of-diacritics}. Further, according to the
spirit of the IPA, it would also be possible to combine more than two base
characters into one tie bar, but this is not possible with Unicode (i.e.~there
is no possibility to draw a tie bar over three of four characters). It is
possible to indicate such larger groups by repeating the tie bar, like in
<a͡o͡u>. If really necessary, we consider this possible, even though the
rendering will never look good.

@@

Although linguists might expect these characters to belong together with the
character in front of them, at least for <ʰ> \textsc{modifier letter small h} at
\uni{02B0} the Unicode Consortium's decision to treat it as a separate character
is also linguistically correct, because according to the IPA it can be used both
for aspiration (more precisely post-aspiration following the base character) and
pre-aspiration (preceding the base character). Note that there exists a mechanism in
Unicode to force separate characters to be combined (namely by using the
\textsc{zero width joiner} at \uni{200D}), but this seems to be a rather
impractical, and probably not an enforceable solution to us.

Further, the IPA tie bar (see Section~\ref{pitfall-tie-bar}) is somewhere in
between these two character classes. The Unicode encoding of this character
belong to the Combining Diacritical Marks, namely either \textsc{combining double
inverted breve} at \uni{0361} or \textsc{combining double breve below} at
\uni{035C}. Such a combining mark is by definition tied to the character in
front, but not the character following it. The Unicode treatment of this
character thus only partly corresponds to the IPA conception, which ideally
would have the tie bar linked both to the character in front and to the
character following.

>>>>

The International Phonetic Alphabet (IPA) is a common standard in linguistics to
transcribe sounds of spoken language into Latin-based characters. Although IPA
is reasonably easily adhered to with pen and paper, it is not trivial to encode
IPA characters electronically. Similar to the previous chapter, this chapter
will discuss various pitfalls with the Unicode encoding of the IPA.\@ We will
specifically refer to the 2005 version of the IPA \citep{IPA2005} and the 7.0 
version of Unicode \citep{Unicode2014}.

@@

Tie bars are a special type of character in the sense that they do not belong to
a segment, but bind two graphemes together. This actually turns out to be rather
different from the Unicode conceptions. The Unicode encoding of this character
belong to the Combining Diacritical Marks, namely either \textsc{combining double
inverted breve} at \uni{0361} or \textsc{combining double breve below} at
\uni{035C}. Such a combining mark is by definition tied to the character in
front, but not the character following it. The Unicode treatment of this
character thus only partly corresponds to the IPA conception, which ideally
would have the tie bar linked both to the character in front and to the
character following. The solution to this is to always use the tie bar in 
combination with

Further, according to the
spirit of the IPA, it would also be possible to combine more than two base
characters into one tie bar, but this is not possible with Unicode (i.e.~there
is no possibility to draw a tie bar over three of four characters). It is
possible to indicate such larger groups by repeating the tie bar, like in
<a͡o͡u>. If really necessary, we consider this possible, even though the
rendering will never look good.

@@

Although linguists might expect these characters to belong together with the
character in front of them, at least for <ʰ> \textsc{modifier letter small h} at
\uni{02B0} the Unicode Consortium's decision to treat it as a separate character
is also linguistically correct, because according to the IPA it can be used both
for aspiration (more precisely post-aspiration following the base character) and
pre-aspiration (preceding the base character). Note that there exists a
mechanism in Unicode to force separate characters to be combined (namely by
using the \textsc{zero width joiner} at \uni{200D}), but to us this seems to be
a rather impractical, and probably not an enforceable solution to the general
problem of complex graphemes. The default combination of Spacing Modifiers with
the preceding character can be specified in orthography profiles (see
Chapter~\ref{orthography-profiles}). Only in case variable directions occur in 
the same transcription (e.g. both pre-aspiration and poast-aspiration occur), 
then each case has to be explicitly marked with a zero-width joiner.

### 2015-06-20 chapters/pitfalls.tex
adding two new unicode pittfalls

Further, the rendering of a glyph is dependent on the font being used. For
example, the Unicode character \textsc{latin small letter g} appears as <g> and
<{\fontspec{Courier}g}> in the Linux Libertine and Courier fonts, respectively,
because their typefaces are designed differently. Furthermore, font face may
change the visual appearance of a character, for example Times New Roman
two-story <{\fontspec{Times New Roman}a}> changes to a single-story a in italics
<\emph{\fontspec{Times New Roman}a}>. This becomes a real problem for some
phonetic typesetting (see Section~\ref{pitfall-ipa-homoglyphs}).

@@

@@

@@

Consider for example the characters and following Unicode code points:
\begin{itemize}
	\def\labelenumi{\arabic{enumi}.} 
	\item[] <Å> \textsc{latin capital letter a with ring above} \uni{00C5} 
	\item[] <Å> \textsc{angstrom sign} \uni{212B}
	\item[] <Å> \textsc{latin capital letter a} \uni{0041}
	+ \textsc{combining ring above} \uni{030A}
\end{itemize}

@@

It is very important to note that any software applications that is Unicode
Standard compliant is free to change the character stream from one
representation to another. This means that a software application may compose,
decompose or reorder characters as its developers desire; as long as the
resultant strings are canonically equivalent to the original. This might lead to
unexpected behavior for users. Various players, like the Unicode Consortium, the
W{\large 3}C, or the TEI recommend NFC in most user-directed situations, and
some software applications that we tested indeed seem to automatically convert
strings into NFC.\footnote{See the summary of various recommendation here:
\url{http://www.win.tue.nl/~aeb/linux/uc/nfc_vs_nfd.html}.} This means in
practice that if a user, for example, enters <a> and < ̀>, i.e.~\textsc{latin
small letter a} at \uni{0061} and \textsc{combining grave accent} at \uni{0300},
this might be automatically converted into <à>, i.e.~\textsc{latin small letter
a with grave} at \uni{00E0}.\footnote{The behavior of software applications can
be quite erratic in this respect. For example, Apple's TextEdit does not do any
conversion on text entry. However, when you copy and paste some text inside the
same document in rich text mode (i.e.~RTF-format), it will be transformed into
NFC on paste. Saving a document does not do any conversion to the glyphs on
screen, but it will save the characters in NFC.}

@@

  \item the \textsc{combining cedilla} at \uni{0327} \newline This diacritic is
        for example attested in the precomposed character <ç> \textsc{latin
        small letter c with cedilla} at \uni{00E7}. This <ç> will thus be
        decomposed in NFC normalization.
  \item the \textsc{combining ogonek} at \uni{0328} \newline This diacritic is
        for example attested in precomposed <ą> \textsc{latin small letter a
        with ogonek} at \uni{0105}. This <ą> will thus be decomposed in NFC
        normalization.
  \item the \textsc{combining horn} at \uni{031B} \newline This diacritic is for
        example attested in precomposed <ơ> \textsc{latin small letter o with
        horn} at \uni{01A1}. This <ơ> will thus be decomposed in NFC
        normalization.

@@

  \item First, there are the combining characters designated as ``combining
        overlay'' in the Unicode Standard, like <\ \ {\large ̴}>
        \textsc{combining tilde overlay} at \uni{0334} or <\ \ {\large ̵}>
        \textsc{combining short stroke overlay} at \uni{0335}. There are many
        characters that look like they are precomposed with such an overlay,
        for example <\charis{ɫ~ᵬ~ᵭ~ᵱ}> or <\charis{ƚ~ɨ~ɉ~ɍ}>, or also the
        example of <ø> given at the start of this section. However, they are 
        not decomposed in NFD normalization.
  \item Second, the same situation also occurs with combining characters
        designated as ``combining hook'', like < {\large ̡}> \textsc{combining
        palatalized hook below} at \uni{0321}. This element seems to occur in
        precomposed characters like <\charis{ᶀ~ᶁ~ᶂ~ᶄ}>. However, they are 
        not decomposed in NFD normalization.

@@

\begin{itemize}
   \item To prevent strange boxes instead of nice glyphs, always install a few
         fonts with a large glyph collection and at least one fall-back font
         (see Section~\ref{pitfall-missing-glyphs}).
   \item Unexpected visual impressions of symbols does not necessarily mean that
         the actual encoding is wrong. It is mostly a problem of faulty
         rendering (see Section~\ref{pitfall-faulty-rendering}).
   \item Do not trust the names of Unicode blocks as a strategy to find specific 
         characters (see Section~\ref{pitfall-blocks})
   \item To ensure consistent encoding of texts, apply Unicode normalization
         (NFC or NFD, see Section~\ref{pitfall-canonical-equivalence}).
   \item To prevent remaining inconsistencies after normalization, for example 
         stemming from homoglyphs (see Section~\ref{pitfall-homoglyphs}) 
         or from missing canonical equivalence in the Unicode Standard
         (see Section~\ref{pitfall-absence-of-equivalence})
         use orthography profiles (see Chapter~\ref{orthography-profiles}).
   \item As a preferred file format, use Unicode Format UTF-8 in 
         Normalization Form Composition (NFC) with LF line endings, 
         but without byte order mark (BOM), whenever possible (see 
         Section~\ref{pitfall-file-formats}). This last nicely cryptic 
         recommendation has T-shirt potential:

>>>>

Further, the rendering of a glyph is dependent on the font being used. For
example, the Unicode character \textsc{latin small letter g} appears as <g> and
<{\fontspec{Courier}g}> in the Linux Libertine and Courier fonts, respectively,
because their typefaces are designed differently. Furthermore, font face may
change the visual appearance of a character, for example Times New Roman
two-story <{\fontspec{Times New Roman}a}> changes to a single-story glyph in italics
<\emph{\fontspec{Times New Roman}a}>. This becomes a real problem for some
phonetic typesetting (see Section~\ref{pitfall-ipa-homoglyphs}).

@@

% ==========================
\section{Pitfall: Characters are not graphemes}
\label{pitfall-characters-are-not-graphemes}
% ==========================

The Unicode Standard is a character encoding system, and not a writing system
encoding system. This most forcefully becomes clear with the notion of grapheme.
From a linguistic point of view, graphemes are the basic building blocks of a
writing system (see Section~\ref{linguistic-terminology}). It is extremely
common, up to the point of being universally attested, that writing systems use
combinations of multiple symbols as a single grapheme, like <sch>, <th> or <ei>.
There is no possibility to encode such complex graphemes using the Unicode Standard.

The Unicode Standard deals with complex graphemes only inasmuch they consist of
base characters with diacritics (see
Section~\ref{pitfall-different-notions-of-diacritics} for a discussion of the
notion of diacritic). The Unicode Standard calls such combination ``grapheme
clusters''. Complex graphemes consisting of multiple base characters,
like <sch>, are called ``tailored grapheme clusters'' in Unicode parlance (see
Section~\ref{the-unicode-approach}).

Inspecting the Unicode Standard, there appear to be special Unicode characters
to ``glue'' together characters into larger tailored grapheme clusters,
specifically the \textsc{zero width joiner} at \uni{200D} and the
\textsc{combining grapheme joiner} at \uni{034F}. However, these characters are
confusingly named (cf.~Section~\ref{pitfall-names}). Both codepoints actually do
not join characters, but explicitly separate them. The zero-width joiner (ZWJ)
can be used to solve special problems related to ordering (called ``collation''
in Unicode parlance). The combining grapheme joiner (CGJ) can be used to
separate characters that are not supposed to form ligatures. To solve the issue
of tailored grapheme clusters, Unicode offers some assistance in the form of the
Unicode Locale Descriptions. However, in the practice of linguistic research,
this is not a real solution. For that reason we propose to use orthography
profiles (see Chapter~\ref{orthography-profiles}).

@@

% ==========================
\section{Pitfall: Names}
\label{pitfall-names}
% ==========================

The names of characters in the Unicode Standard are sometimes misnomers and
should not be misinterpreted as definitions. For example, the \textsc{combining
grapheme joiner} at \uni{034F} does not join characters into larger graphemes
(see Section~\ref{pitfall-characters-are-not-graphemes}) and the \textsc{latin
letter retroflex click} \uni{01C3} is actually not the IPA symbol for a
retroflex click, but for an alveolar click (see
Section~\ref{pitfall-ipa-homoglyphs}). In a sense, these names can be seen as
``errors.'' However, it is probably better to realize that such names are just
convenience labels that are not going to be changed. Just like the block names
(Section~\ref{pitfall-blocks}), the character names are often helpful, but they
are not supposed to be definitions.

The actual intended ``meaning'' of a Unicode codepoint is a combination of the
name, the block and the character properties (see
Section~\ref{the-unicode-approach}). Further details about the underlying intentions 
with which a character should be used
are only accessible by perusing the actual decisions of the Unicode Consortium.
All proposals, discussions and decisions of the Unicode Consortium are publicly
available. Unfortunately there is not (yet) any way to easily find everything
that is ever proposed, discussed and decided in relation to a specific
codepoint of interest, so many of the details are often somewhat
hidden.\footnote{All proposals and other documents that are the basis of Unicode
decisions are avaialbe at \url{http://www.unicode.org/L2/all-docs.html}. The
actual decisions that make up the Unicode Standard are documented in the minutes
of the Unicode Technical Committee, available at
\url{http://www.unicode.org/consortium/utc-minutes.html}.}

@@

Consider for example the characters and following Unicode code points:
\begin{itemize}
	\def\labelenumi{\arabic{enumi}.} 
	\item <Å> \textsc{latin capital letter a with ring above} \uni{00C5} 
	\item <Å> \textsc{angstrom sign} \uni{212B}
	\item <Å> \textsc{latin capital letter a} \uni{0041}
	+ \textsc{combining ring above} \uni{030A}
\end{itemize}

@@

It is very important to note that any software applications that is Unicode
Standard compliant is free to change the character stream from one
representation to another. This means that a software application may compose,
decompose or reorder characters as its developers desire; as long as the
resultant strings are canonically equivalent to the original. This might lead to
unexpected behavior for users. Various players, like the Unicode Consortium, the
W{\large 3}C, or the TEI recommend NFC in most user-directed situations, and
some software applications that we tested indeed seem to automatically convert
strings into NFC.\footnote{See the summary of various recommendation here:
\url{http://www.win.tue.nl/~aeb/linux/uc/nfc_vs_nfd.html}.} This means in
practice that if a user, for example, enters <a> and <\dia{0300}>, i.e.~\textsc{latin
small letter a} at \uni{0061} and \textsc{combining grave accent} at \uni{0300},
this might be automatically converted into <à>, i.e.~\textsc{latin small letter
a with grave} at \uni{00E0}.\footnote{The behavior of software applications can
be quite erratic in this respect. For example, Apple's TextEdit does not do any
conversion on text entry. However, when you copy and paste some text inside the
same document in rich text mode (i.e.~RTF-format), it will be transformed into
NFC on paste. Saving a document does not do any conversion to the glyphs on
screen, but it will save the characters in NFC.}

@@

  \item <\dia{0327}>: the \textsc{combining cedilla} at \uni{0327} \newline 
        This diacritic is
        for example attested in the precomposed character <ç> \textsc{latin
        small letter c with cedilla} at \uni{00E7}. This <ç> will thus be
        decomposed in NFC normalization.
  \item <\dia{0328}>: the \textsc{combining ogonek} at \uni{0328} \newline 
        This diacritic is
        for example attested in precomposed <ą> \textsc{latin small letter a
        with ogonek} at \uni{0105}. This <ą> will thus be decomposed in NFC
        normalization.
  \item <\dia{031B}>: the \textsc{combining horn} at \uni{031B} \newline 
        This diacritic is for
        example attested in precomposed <ơ> \textsc{latin small letter o with
        horn} at \uni{01A1}. This <ơ> will thus be decomposed in NFC
        normalization.

@@

  \item First, there are the combining characters designated as ``combining
        overlay'' in the Unicode Standard, like <\dia{0334}>
        \textsc{combining tilde overlay} at \uni{0334} or <\dia{0335}>
        \textsc{combining short stroke overlay} at \uni{0335}. There are many
        characters that look like they are precomposed with such an overlay,
        for example <\charis{ɫ~ᵬ~ᵭ~ᵱ}> or <\charis{ƚ~ɨ~ɉ~ɍ}>, or also the
        example of <ø> given at the start of this section. However, they are 
        not decomposed in NFD normalization.
  \item Second, the same situation also occurs with combining characters
        designated as ``combining hook'', like 
        <{\fontspec{CharisSIL}{\large ◌}}\symbol{"0321}> \textsc{combining
        palatalized hook below} at \uni{0321}. This element seems to occur in
        precomposed characters like <\charis{ᶀ~ᶁ~ᶂ~ᶄ}>. However, they are 
        not decomposed in NFD normalization.

@@

\begin{itemize}
   \item To prevent strange boxes instead of nice glyphs, always install a few
         fonts with a large glyph collection and at least one fall-back font
         (see Section~\ref{pitfall-missing-glyphs}).
   \item Unexpected visual impressions of symbols does not necessarily mean that
         the actual encoding is wrong. It is mostly a problem of faulty
         rendering (see Section~\ref{pitfall-faulty-rendering}).
   \item Do not trust the names of codepoints as a definition of the character
         (see Section~\ref{pitfall-names}). Also do not trust Unicode blocks as
         a strategy to find specific characters (see
         Section~\ref{pitfall-blocks})
   \item To ensure consistent encoding of texts, apply Unicode normalization
         (NFC or NFD, see Section~\ref{pitfall-canonical-equivalence}).
   \item To prevent remaining inconsistencies after normalization, for example 
         stemming from homoglyphs (see Section~\ref{pitfall-homoglyphs}) 
         or from missing canonical equivalence in the Unicode Standard
         (see Section~\ref{pitfall-absence-of-equivalence})
         use orthography profiles (see Chapter~\ref{orthography-profiles}).
   \item To deal with ``tailored'' grapheme clusters
         (Section~\ref{pitfall-characters-are-not-graphemes}), use Unicode Locale 
         Descriptions, or orthography profiles 
         (see Chapter~\ref{orthography-profiles}).
   \item As a preferred file format, use Unicode Format UTF-8 in 
         Normalization Form Composition (NFC) with LF line endings, 
         but without byte order mark (BOM), whenever possible (see 
         Section~\ref{pitfall-file-formats}). This last nicely cryptic 
         recommendation has T-shirt potential:
